\Large{All in the family:\\Anna and Lisa Hahner's finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-03 paper/paper.tex
paper updates on friday later

We now consider the following: if we take into consideration the
Hahner twin's similarities as runners as well as some natural
variation in finishing a marathon, what is the probability that the
twins finished at roughly the same time and/or sequentially?  To
answer this question we need to know the counterfactual distribution
of potential marathon finishing times that would have occurred had the
Hahner twins independently and repeatedly run the Rio marathon over
again.  Holding the twin's abilities constant, this would establish
the set of potential outcomes that would have occurred as result of
natural variation in the finishing the marathon.  We could use this
distribution of outcomes to determine the likelihood that a
simultaneous finish by the Hanhers, or at least a near-simultaneous
finish, occurred by chance alone.  If the twins rarely finish the
marathon together in such a counterfactual world, then one might be
skeptical that their observed finish occurred without
coordination---as they in fact claimed.

Unfortunately for us but nonetheless fortunately for the race's
participants, it is not possible to rerun the women's Olympic marathon
in Rio to establish a counterfactual distribution of potential race
outcomes for the Hahner twins.  However, we can attempt to simulate
this distribution by estimating the distribution of every other
runner's finishing time, conditioned on prior abilities, and drawing
from that distribution in order to observe the likelihood that Lisa
and Anna Hahner finish simultaneously.

To estimate the conditional distribution of Rio runner's result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of her personal best time
$X_i$:

$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right).$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. We
estimate $\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least
squares and our set of 133 finishing runners.  For a simulated
marathon, we then draw a runner's time from an estimated distribution,
conditioned on the runner's personal best outcome.  Once a race is
simulated, we locate Anna and Lisa Hahner and record both the time
between their finishes and the difference in their ranks.  We then
simulate a new race---drawing a new set of finishing times---and
record the same quantities.  In particular, the steps of the
simulation are as follows.

\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time ($Y_i$) based
  on her pre-Olympic personal best time ($X_i$) and her best time
  squared.
\item Extract coefficient vector, estimated covariance matrix, and
  estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the estimated coefficient vector and covariance equal to
  the estimated covariance matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item For each runner, predict the final result by combining the randomly
  generated beta coefficients and error terms, $\tilde{y} = \tilde{\beta}_0 +
  \tilde{\beta}_1\,X_i + \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the race with a probability equal to
  the observed fraction of marathoners who did not finish the
  marathon.
\item Calculate the difference in time and difference in ranking
  between Anna Hahner and Lisa Hahner, assuming both finished the
  race.
\item Repeat above steps 10,000 times.
\end{enumerate}

Before considering results, DNF biz...

Figure \ref{fig:simdiff} contains two histograms of differences
between Anna and Lisa's finishes.  Figure
\ref{fig:simulatedfinishtimes} shows the distribution of absolute
differences in time where differences are grouped in 30 second bins
along the horizontal axis of the histogram.  The count for each bin is
denoted by the vertical length of the bars.  Figure
\ref{fig:simulatedranks} shows the distribution of the absolute
differences in Hahner twin rankings.  Differences are grouped as
single units ranging from no runner between the twins to nearly 120
runners between them.

@@

The results plotted in the histograms question the credibility of Anna
and Lisa's story.  The results suggest that a simultaneous finish
would be very rare if Anna and Lisa had run independently of each
other.  For example, in less than 300 of the 10,000 simulated races
did Anna and Lisa finished within 30 seconds of each other .  And they
finished in consecutive rank in less than 175 of the 10,000 races.
Therefore, the close finish that we observed, where Anna and Lisa
crossed the finish line one-after-the-other, would be very unlikely if
we were to believe the twins raced independently of each other.

The results holds even if you replace their personal best time with
their time halfway through the race.  While this reduces the variation
in the predicted outcome of each runner and, therefore, reduces the
expected distance between Anna and Lisa, it is still quite rare for
Anna and Lisa to finish simultaneously or consecutively.

>>>>

We now consider the following question: if we take into consideration
the Hahner twin's similarities as runners as well as natural variation
in marathon finishing times, what is the probability that the twins
would finish the Rio race at roughly the same time and/or
sequentially?  To answer this question we need to know the
counterfactual distribution of potential marathon finishing times that
would have occurred had the Hahner twins independently (in particular,
of each other) and repeatedly run the Rio marathon, holding constant
marathon conditions, the abilities of other runners, and so forth.
Access to such a distribution would establish the set of potential
outcomes that could have occurred on August 14, 2016, and we could in
principle use this distribution to determine the likelihood that a
simultaneous finish by the Hahners, or at least a near-simultaneous
finish, occurred by chance alone.  If these twins rarely finish the
marathon together in such a counterfactual world, then one might be
skeptical that their observed finish occurred without
coordination---as they in fact claimed.

Unfortunately for us but nonetheless fortunately for the race's
participants, it is not possible to rerun the women's Olympic marathon
in Rio to establish a distribution of potential race outcomes for the
Hahner twins.  However, we can attempt to simulate this distribution
by estimating the distribution of every other runner's finishing time,
conditional on prior abilities, and drawing from that distribution in
order to observe the likelihood that, for example, Lisa and Anna
Hahner finished simultaneously.

To estimate the conditional distribution of each Rio final results, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of her personal best time
$X_i$:

$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + e_i, \sigma\right).$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution.
Personal best times are all prior to the Rio marathon.  We estimate
$\beta_1$ , $\beta_2$, and $\sigma$ using ordinary least squares and
all finishing runners except for the twins and triplets.  For a
simulated marathon, we then draw a runner's time from an estimated
distribution, conditioned on the runner's personal best marathon time.
Once a race is simulated for all runners, twins and triplets included,
we locate Anna and Lisa Hahner in our set of simulated results and
record both the time between their simulated finishes and the
difference in their simulated ranks.  We then simulate a new
race---drawing a new set of finishing times---and record the same
quantities.  In particular, the steps of the simulation are as
follows.

\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her best time squared.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and $\tilde{\beta}_2$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
\item Calculate the difference in time and difference in ranking
  between Anna Hahner and Lisa Hahner, assuming both finished the
  race.
\item Repeat above steps 10,000 times.
\end{enumerate}

%% Before considering results, DNF biz...

From our simulations we can generate intervals which describe the
extent of the variability in racer finishing times.  In 95\% of the
simulations in which Anna Hahner completed the marathon, her finishing
time was between 9,480 seconds and 10,703 seconds.  This is consistent
with Anna's observed finishing time, which was 9,932 seconds.  In
fact, our simulations estimate that Anna should have finished the Rio
marathon in 10,092 seconds, which is slightly more than two and
one-half minutes slower than her actual time.  Lisa Hahner's
corresponding interval is 9,427 to 10,619 with an observed finishing
time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times.  And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets).
According to the our simulations, then, neither Hahner twin ran
appreciably slowly, conditional on personal best times prior to the
Rio Olympics.

% 95% simulated interval for Anna Hahner finishing time in seconds: 9480 10703 
% 95% simulated interval for Lisa Hahner finishing time in seconds: 9427 10619

% > df$FINAL[df$ATHLETE == "Anna Hahner"]
% [1] 9932
% > df$FINAL[df$ATHLETE == "Lisa Hahner"]
% [1] 9933

With an eye on the matter of simultaneous finishing, Figure
\ref{fig:simdiff} contains two histograms based on our simulated race
results. Figure \ref{fig:simulatedfinishtimes} is a histogram which
shows the distribution of absolute differences in Hahner twin
finishing time where differences are grouped in 30 second bins; counts
for the various bins are denoted by the vertical lengths of the bars.
Figure \ref{fig:simulatedranks} is similar but depicts the
distribution of the absolute differences in Hahner twin rankings.
Differences are grouped as single units ranging from no runner between
the twins to nearly 120 runners between them.

@@

The results plotted in the histograms raise questions about the
credibility of Anna and Lisa Hahner's story.  In particular, the
histogram suggest that a simultaneous finish would be very rare if
Anna and Lisa had run independently of each other.  For example, in
fewer than 300 of the 10,000 simulated races did Anna and Lisa
finished within 30 seconds of each other.  Moreover, they finished in
consecutive rank in fewer than 175 of the 10,000 races.  Therefore,
the close finish that was observed, where Anna and Lisa crossed the
finish line one-after-the-other, would be highly unlikely if the two
German twins had raced independently of each other.

Earlier we noted that personal best times might arguably be poor
measures of runner abilities.  With this in mind, we repeated our
simulations but replaced personal best times with half marathon
splits.  See Figure \ref{fig:simdiffhalf}.  While this change reduces
the variation in the predicted outcome of each runner and therefore,
reduces the expected distance between Anna and Lisa Hahner, even with
half marathon split as a measure of ability it is still quite rare for
Anna and Lisa to finish simultaneously or consecutively.
